\section{Method}
\label{sec:method}

We extend the paragraph-level graph neural network of \citet{ilpcsr} with three modifications informed by the analysis in \S\ref{sec:analysis}. We first describe the base architecture, then the extensions.

\subsection{Base Architecture: Para-GNN}
\label{sec:base}

We adapt Para-GNN \citep{ilpcsr} for multilingual statute retrieval. For each query-candidate pair, a heterogeneous graph is constructed with four node types: (i) query document nodes, (ii) candidate document nodes, (iii) query paragraph nodes (one per sentence), and (iv) candidate paragraph nodes (one per statute clause). Each paragraph node is connected to its parent document node via a directed edge. Node features are sentence-level embeddings; edge features are rhetorical role embeddings (``NONE'' for statutes). Document node features are the mean of their paragraph embeddings.

A two-layer Edge-Updated Graph Attention Network (EUGAT) processes the graph, updating node representations through learned attention over neighbors and edge features, with residual connections. The final score for a query-candidate pair is:
\begin{equation}
\text{score} = \alpha \cdot \hat{s}_{\text{GNN}} + (1-\alpha) \cdot s_{\text{BM25}}
\label{eq:score}
\end{equation}
where $\hat{s}_{\text{GNN}}$ is the Z-normalized GNN score and $\alpha$ is a blending weight.

We replace the original English-only encoder (all-mpnet-base-v2, 768d) with BGE-M3 \citep{bgem3}, a multilingual encoder (1024d) supporting Indonesian, French, Chinese, and English. All other components (EUGAT layers, training loss, negative sampling) follow \citet{ilpcsr}.

\subsection{Post-Training Alpha Grid Search}
\label{sec:alpha}

In \citet{ilpcsr}, $\alpha$ is learned end-to-end via a feed-forward network. We observe that this learned $\alpha$ converges to approximately 0.97 regardless of dataset, effectively ignoring BM25 (\S\ref{sec:analysis}). However, the optimal $\alpha$ varies from 0.7 to 0.9 depending on the level of lexical overlap in each dataset.

We decouple signal blending from GNN training. After training, we compute the raw GNN scores and BM25 scores for all test query-candidate pairs, then sweep $\alpha \in \{0.0, 0.1, \ldots, 1.0\}$ and select the value that maximizes MRR@10. This takes less than one second as it involves only arithmetic on pre-computed score matrices.

\subsection{Structural Node Features}
\label{sec:structural}

In the base Para-GNN graph, each statute node is connected only to its own clause nodes. Statute nodes are not connected to each other, even when the corresponding articles belong to the same legal instrument. Our analysis shows that co-relevant statutes cluster within the same instrument (85.2\% share the same book) and are structurally close (median 18 articles apart), while hard negatives are distant (median 297). We encode this structural information directly into node features, allowing the GNN to learn which structural patterns predict co-relevance.

\noindent\textbf{Act Membership.} Each statute belongs to a legal instrument (e.g., an act, code, or law). We extract the instrument name from the statute title and encode it as a deterministic hash vector $\mathbf{a} \in \mathbb{R}^{64}$. The hash function maps each instrument name to a unit-norm pseudo-random vector using a fixed seed. Statutes from the same instrument receive identical vectors; statutes from different instruments receive near-orthogonal vectors, as high-dimensional random unit vectors have expected dot product approaching zero.

\noindent\textbf{Positional Rank.} Within each instrument, statutes are ordered by their index of appearance in the corpus, normalized to $[0, 1]$. We encode this rank using sinusoidal positional encoding $\mathbf{p} \in \mathbb{R}^{32}$ at multiple frequencies, following \citet{transformer}. This captures structural proximity at different scales: adjacent articles produce similar encodings at all frequencies, while articles in the same chapter share low-frequency components but differ at high frequencies.

\noindent\textbf{Query Nodes.} Query nodes do not belong to any legal instrument. They receive a dedicated hash vector (distinct from all instrument hashes) and a fixed midpoint position encoding ($r = 0.5$), providing the GNN with a stable structural signature for query nodes.

\noindent\textbf{Projection.} For each node with semantic embedding $\mathbf{e}_i$, act hash $\mathbf{a}_i$, and positional encoding $\mathbf{p}_i$, we compute:
\begin{equation}
\mathbf{x}_i = W \cdot \text{LayerNorm}([\mathbf{e}_i \| \mathbf{a}_i \| \mathbf{p}_i])
\label{eq:struct-proj}
\end{equation}
where $W \in \mathbb{R}^{1024 \times 1120}$ is a learned projection matrix. The linear projection preserves the geometric orthogonality of act hashes. The projected features $\mathbf{x}_i$ are then passed to EUGAT in place of the raw semantic embeddings.

With structural features concatenated into node representations, the EUGAT attention scores between candidate statutes incorporate both semantic similarity and structural relatedness. Statutes from the same instrument with similar positions receive higher attention weights, enabling the GNN to propagate relevance signal between structurally related candidates without requiring explicit inter-statute edges.

\subsection{Baselines}

We compare against four baselines using the same evaluation protocol.

\noindent\textbf{BM25:} Okapi BM25 with $k_1=1.5$, $b=0.75$. Stemming disabled for Indonesian; jieba segmentation for Chinese (details in App.~\ref{app:implementation}).

\noindent\textbf{Dense (BGE-M3):} Cosine similarity over BGE-M3 dense vectors.

\noindent\textbf{GAR \citep{gar}:} BM25 pool (top 200) re-ranked with a cross-encoder scorer and expanded via corpus graph traversal.

\noindent\textbf{JNLP Stage~1 \citep{jnlp}:} CatBoost classifier on element-wise product features of BGE-M3 query and document embeddings.

\noindent\textbf{Neural Sparse \citep{opensearch_sparse}:} Multilingual learned sparse retrieval. Documents are encoded via an MLM forward pass with $\log(1 + \text{ReLU}(\cdot))$ activation and max-ratio pruning; queries use inference-free IDF-based term weights. Relevance is scored by sparse dot product over the shared vocabulary space.

\noindent\textbf{MILCO \citep{milco}:} Multilingual learned sparse retrieval via language-agnostic projection into English lexical space. A multilingual connector maps hidden states to an English MLM head, and a LexEcho head augments the English representation with source-language token weights to preserve entities that lack English equivalents. Scored by sparse dot product.

\subsection{Agentic Retrieval}
\label{sec:agentic}

We evaluate an LLM-orchestrated retrieval agent that replaces single-pass scoring with an iterative observe-reason-act loop. The agent receives a query and accesses the corpus through four tools: \texttt{search\_corpus} (hybrid BM25 + StructGNN retrieval), \texttt{grep\_corpus} (lexical pattern matching), \texttt{read\_document} (full article text), and \texttt{prune\_chunks} (remove low-relevance passages from context).

The agent follows a structured reasoning protocol. In the first turn, it performs \emph{frame enumeration}: identifying the legal interests at stake and enumerating doctrinal frames (bodies of law) that could govern each interest. Subsequent turns issue targeted searches derived from (frame, doctrine) pairs rather than surface query terms. Before concluding, the agent produces a coverage table verifying that every enumerated frame has been either covered (provision found) or explicitly rejected.

The agent operates with a maximum of 10 turns. A similarity guard (cosine threshold 0.92) prevents near-duplicate provisions from entering the selection set, and a coverage gate ensures all enumerated frames are addressed before the agent terminates.

\subsection{Evaluation Protocol}

We report MRR@10, Recall@10, and Hit Rate. Each method is evaluated on both KUHPerdata variants (shared relevance judgments) and on BSARD and STARD at their native query levels. All datasets use the same train and test split protocol.
