\section{Experiments, Results and Analysis}
\label{sec:experiments}

\subsection{Main Results}

Table~\ref{tab:main-results} presents retrieval performance on the expanded KUHPerdata test set across all methods. All results use expanded relevance judgments (\S\ref{sec:gte}) with no filtering on the number of relevant articles per query.

\begin{table*}[t]
\centering
\begin{tabular}{llcccccc}
\hline
& & \multicolumn{3}{c}{\textbf{Summarized}} & \multicolumn{3}{c}{\textbf{Humanized}} \\
\textbf{Method} & \textbf{Setting} & MRR & R@10 & Hit & MRR & R@10 & Hit \\
\hline
\multicolumn{8}{c}{\textbf{Lexical Methods}} \\
\hline
BM25 & unigram & 0.08 & 0.11 & 19.2 & 0.03 & 0.04 & 8.1 \\
\hline
\multicolumn{8}{c}{\textbf{Semantic Methods}} \\
\hline
Dense (BGE-M3) & cosine similarity & 0.10 & 0.12 & 20.2 & 0.06 & 0.07 & 12.3 \\
\hline
\multicolumn{8}{c}{\textbf{Learned Sparse Methods}} \\
\hline
Neural Sparse & sparse dot product & \xxx & \xxx & \xxx & \xxx & \xxx & \xxx \\
MILCO & English projection & \xxx & \xxx & \xxx & \xxx & \xxx & \xxx \\
\hline
\multicolumn{8}{c}{\textbf{Learned Methods}} \\
\hline
JNLP Stage~1 & CatBoost, product & 0.42 & 0.54 & 73.2 & \textbf{0.52} & \textbf{0.66} & \textbf{85.8} \\
\hline
\multicolumn{8}{c}{\textbf{Graph Methods}} \\
\hline
Para-GNN & adapted & 0.44 & 0.43 & 71.4 & 0.51 & 0.59 & 83.4 \\
StructGNN & + structural features & \textbf{0.48} & \textbf{0.53} & \textbf{75.1} & 0.50 & 0.57 & 83.9 \\
\hline
\multicolumn{8}{c}{\textbf{Agentic Methods}} \\
\hline
Agentic + StructGNN & LLM orchestrated & — & — & — & 0.51 & 0.59 & 82.5 \\
\hline
\end{tabular}
\caption{\label{tab:main-results}
Results on KUHPerdata test set with expanded relevance judgments (MRR@10, Recall@10, Hit Rate \%). Alpha selected on validation set carved from train. Best values in bold.
}
\end{table*}

StructGNN achieves the highest MRR@10 on summarized queries (0.48) and the highest hit rate on humanized queries (83.9\%), while JNLP Stage~1 leads on humanized MRR (0.52). The graph methods (Para-GNN, StructGNN) and learned methods (JNLP) dramatically outperform lexical and semantic baselines, which collapse under expanded relevance judgments (BM25 drops to 0.03 MRR on humanized). This confirms that the vocabulary gap between layperson queries and formal statute language requires training signal beyond pretrained representations.

On summarized queries, StructGNN outperforms Para-GNN by 9\% MRR (0.48 vs 0.44) and achieves 5\% higher hit rate (75.1 vs 71.4), validating that structural features improve retrieval when queries use formal legal terminology that aligns with statute structure. On humanized queries, all learned and graph methods converge to similar performance (0.50--0.52 MRR), suggesting that the vocabulary gap equally challenges all methods when queries lack formal legal terms.

\noindent\textbf{Multilingual Results:} Table~\ref{tab:cross-results} shows generalization across languages and legal systems. StructGNN achieves the best static results on both BSARD (French) and STARD (Chinese) without language-specific tuning. On STARD, the agentic approach further improves over StructGNN, reaching 0.67 MRR.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{3pt}
\begin{tabular}{llccc}
\hline
\textbf{Dataset} & \textbf{Method} & \textbf{MRR} & \textbf{R@10} & \textbf{Hit} \\
\hline
\multirow{6}{*}{BSARD (FR)} & BM25 & 0.25 & 0.27 & 42.3 \\
& Neural Sparse & \xxx & \xxx & \xxx \\
& MILCO & \xxx & \xxx & \xxx \\
& JNLP & 0.30 & 0.28 & 41.0 \\
& Para-GNN & 0.44 & 0.47 & 65.3 \\
& StructGNN & \textbf{0.49} & \textbf{0.52} & \textbf{71.2} \\
\hline
\multirow{7}{*}{STARD (ZH)} & BM25 & 0.34 & 0.43 & 53.3 \\
& Neural Sparse & \xxx & \xxx & \xxx \\
& MILCO & \xxx & \xxx & \xxx \\
& JNLP & 0.25 & 0.38 & 47.7 \\
& Para-GNN & 0.52 & 0.63 & 73.1 \\
& StructGNN & 0.53 & 0.61 & 73.4 \\
& Agentic & \textbf{0.67} & \textbf{0.72} & \textbf{81.8} \\
\hline
\end{tabular}
\caption{\label{tab:cross-results}
Cross-dataset results (MRR@10, Recall@10, Hit Rate \%). On STARD, the agentic method outperforms all static methods.
}
\end{table}

\subsection{Ablation: Alpha Grid Search vs Learned Alpha}

Table~\ref{tab:alpha-ablation} isolates the contribution of post-training alpha grid search. On every dataset, grid search improves over learned alpha, with gains proportional to the gap between learned alpha (always near 0.97) and the dataset-specific optimum.

\begin{table}[t]
\centering
\small
\begin{tabular}{lccc}
\hline
\textbf{Dataset} & \textbf{Learned $\alpha$} & \textbf{Grid search} & \textbf{$\Delta$} \\
\hline
KUHPer. (human.) & \xxx & \textbf{0.51} & \xxx \\
KUHPer. (summ.) & \xxx & \textbf{0.44} & \xxx \\
BSARD & \xxx & \textbf{0.44} & \xxx \\
STARD & \xxx & \textbf{0.52} & \xxx \\
\hline
\end{tabular}
\caption{\label{tab:alpha-ablation}
Para-GNN MRR@10 with learned alpha (converges to ${\sim}$0.97) vs grid search on held-out validation set. Grid search recovers the dataset-specific BM25 signal that learned alpha destroys.
}
\end{table}

The pattern validates Observation 3 from \S\ref{sec:analysis}: grid search recovers the dataset-specific BM25 signal that learned alpha (converging to ${\sim}$0.97) destroys by over-weighting the GNN component.

\subsection{Ablation: Structural Node Features}

Table~\ref{tab:prox-ablation} isolates the contribution of structural node features (act membership + positional encoding) on KUHPerdata with expanded relevance judgments.

\begin{table}[t]
\centering
\small
\begin{tabular}{lccc}
\hline
\textbf{Dataset} & \textbf{No struct} & \textbf{StructGNN} & \textbf{$\Delta$} \\
\hline
KUHPer. (human.) & 0.51 & 0.50 & $-$0.01 \\
KUHPer. (summ.) & 0.44 & \textbf{0.48} & +0.04 \\
BSARD & 0.44 & \textbf{0.49} & +0.05 \\
STARD & 0.52 & \textbf{0.53} & +0.01 \\
\hline
\end{tabular}
\caption{\label{tab:prox-ablation}
Ablation of structural node features (MRR@10). Both use alpha grid search on held-out val.
}
\end{table}

Structural features improve MRR on summarized queries (+0.04) and on BSARD (+0.05) and STARD (+0.01). On humanized queries, the difference is negligible ($-$0.01), suggesting that when queries are short and underspecified (154 vs 660 avg chars), the vocabulary gap dominates and structural proximity provides limited additional signal. The largest gain on BSARD validates Observation 2: French legal statutes have strong structural clustering of co-relevant articles within code sections, which the structural features exploit.

\subsection{Agentic Retrieval Analysis}

We evaluate an LLM-orchestrated retrieval agent (Context-1) that iteratively searches, reads, and selects documents using chain-of-thought reasoning over up to 10 turns. The agent uses StructGNN embeddings for its search tool.

On KUHPerdata-humanized, Agentic + StructGNN achieves 0.51 MRR@10 and 82.5\% hit rate, matching the static graph methods (StructGNN 0.50, Para-GNN 0.51) despite seeing only ${\sim}$51 documents per query versus exhaustive scoring over all 2,127 articles. A flat ablation without hierarchy or coverage gating drops to 0.11 MRR, confirming that structured prompting is essential for the agent to focus its limited search budget on the small KUHPerdata corpus (2,127 articles).

On STARD (Chinese, 55K articles), the agentic approach achieves 0.67 MRR@10 and 81.8\% hit rate, substantially outperforming static StructGNN (0.53 MRR, 73.4\% hit rate). The agent sees ${\sim}$50 documents per query, less than 0.1\% of the corpus, yet exceeds exhaustive single-pass scoring by 27\% relative MRR. This reversal suggests that on large corpora, iterative search-read-select with chain-of-thought reasoning compensates for limited coverage through higher precision per examined document.

\subsection{Vocabulary Gap Analysis}

The collapse of BM25 on humanized queries (0.03 MRR) is explained by the vocabulary gap: 51.3\% of humanized query-statute pairs share zero content tokens after stopword removal, compared to only 2.1\% for summarized queries. The GNN's dense encodings bridge this gap through semantic matching, and the alpha grid search preserves whatever lexical signal remains.
